Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-30 ThesisJP/Background/background.tex
Finished fixing text based on comments from Ines.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing a ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short impulse consist of many frequency components. A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth. An ultrasound system has normally a relative bandwidth between 50 \% and 100 \%. This corresponds to transmitting a pulse which is between one or two oscillations when entering tissue. In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one A-mode image is after this process given by:
\begin{align}
N = \frac{2r}{c}B,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse). Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10. Newer versions of PCI Express which supports even higher transfer rates have also been developed.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound imaging systems generates an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan. The data rate from a 2D scan is "only" in order of giga bytes per second prior to IQ-demodulation. To handle these amount of data ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computationally power of CPUs have increased, ASICs, e.g. for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is the beamformer \cite{Thomenius}, whose job is to steer and focus the ultrasound beam in space. This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefor very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

>>>>

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of complex element data every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express which supports even higher transfer rates have also recently been introduced.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound scanners generate an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan. Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner. The data rate of a 2D scan is "only" in order of giga bytes per second. To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}. As seen from (\ref{eq:das}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefore very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up for new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
